TADCompare: an R package for differential analysis and characterization of Topologically Associated Domains

### 2019-09-29 manuscript/manuscript.Rmd
minor discussion polish

Our method relies on the ability to link the phenotype-associated genetic variants detected in genome-wide association studies to genes thought to be affected by them. Typically, genes are linked to genetic variants by proximity (nearest gene), a strategy adopted by the NHGRI-EBI GWAS catalog [@MacArthur:2017aa]. <!--Numerous tools have been created to associate genetic variants with genes (MAGENTA [@Segre:2010aa], VEGAS [@Liu:2010aa], SKAT [@Ionita-Laza:2013aa], to name a few, reviewed in [@Wang:2011ab]).--> However, the majority of phenotype-associated genetic variants are located in regulatory elements outside of protein-coding regions [@Khurana:2013aa]. <!--classical genetic studies of translocations and deletions have shown that distal regions, located at tens to hundreds of kilobases from a gene, affect transcription, indicating that regulatory elements can exert their effect over large genomic distances [@Kleinjan:2005aa].-->The pilot project by ENCODE showed that fewer than 10% of all interactions between TSSs and regulatory regions involved the closest TSS by linear distance [@Sanyal:2012aa], questioning the validity of the nearest gene mapping strategy. The development of chromatin conformation capture technologies now allows for understanding interactions among genomic regions on a genome-wide scale [@Lieberman-Aiden:2009aa; @Zhang:2013aa], helping to prioritize disease-associated gene-variant associations [@Pan:2018aa]. Consequently, tools are being developed to consider long-distance interactions when mapping SNPs to genes (3DSNP [@Lu:2017aa], FUMA [@Watanabe:2017aa], PINES [@Bodea2016], HUGIn [@Martin:2017aa]), and studies redefining disease-associated genes using long-distance interactions started to emerge [@Fang:2019aa]. Our future goal includes considering long-distance variant-gene interactions in defining phenotype-associated gene sets and comparing their network properties.

>>>>

Our method relies on the ability to link the phenotype-associated genetic variants detected in genome-wide association studies to genes thought to be affected by them. Typically, genes are linked to genetic variants by proximity (nearest gene), a strategy adopted by the NHGRI-EBI GWAS catalog [@MacArthur:2017aa]. <!--Numerous tools have been created to associate genetic variants with genes (MAGENTA [@Segre:2010aa], VEGAS [@Liu:2010aa], SKAT [@Ionita-Laza:2013aa], to name a few, reviewed in [@Wang:2011ab]).--> However, >88% of phenotype-associated genetic variants are located in regulatory elements outside of protein-coding regions [@Hindorff:2009aa]. <!--classical genetic studies of translocations and deletions have shown that distal regions, located at tens to hundreds of kilobases from a gene, affect transcription, indicating that regulatory elements can exert their effect over large genomic distances [@Kleinjan:2005aa].-->The pilot project by ENCODE showed that fewer than 10% of all interactions between TSSs and regulatory regions involved the closest TSS by linear distance [@Sanyal:2012aa], questioning the validity of the nearest gene mapping strategy. The development of chromatin conformation capture technologies now allows for understanding interactions among genomic regions on a genome-wide scale [@Lieberman-Aiden:2009aa; @Zhang:2013aa], helping to prioritize disease-associated gene-variant associations [@Pan:2018aa]. Consequently, tools are being developed to consider long-distance interactions when mapping SNPs to genes (3DSNP [@Lu:2017aa], FUMA [@Watanabe:2017aa], PINES [@Bodea2016], HUGIn [@Martin:2017aa]), and studies redefining disease-associated genes using long-distance interactions started to emerge [@Fang:2019aa]. Our future goal includes considering long-distance variant-gene interactions in defining phenotype-associated gene sets and comparing their network properties.
